Permutation Pattern matching in\\ $(213,231)$-avoiding permutations

### 2016-10-13 article.tex
revision

We focus in this paper on pattern matching issues for
$(213,231)$-avoiding permutations that we call wedge permutations.
(\emph{i.e.}, those permutations that avoid both $213$ and $231$).
%The number of $n$-wedge permutations is
%$t_0 = 1$ for $n = 0$ and
%$t_n =2^{n-1}$ for $n\geq 1$ \cite{Simion:Schmidt:EJC:1985},
%as they are in bijection with binary word of size $n-1$.
%On an individual basis,
%the permutations that do not have an occurrence of the permutation pattern $231$
%are exactly the \emph{stack-sortable permutations} and they are counted by
%the Catalan numbers \cite{Knuth:1997:ACP:260999}.
%	A stack-sortable permutation is a permutation whose elements may be sorted by
%	an algorithm whose internal storage is limited to a single stack data structure.
%	As for $213$, it is well-known that
% 	if $\pi = \pi_1\pi\,\ldots\,\pi_n$ avoids $132$, then its complement
% 	$\pi' = (n+1-\pi_1)(n+1-\pi_2)\,\ldots\,(n+1-\pi_n)$ avoids $312$, and
% 	the reverse of $\pi'$ avoids $213$.
% From a combinatorial point of view,
% B\'ona \cite{Bona:ElJC:2012}
% showed the rather surprising fact that the cumulative number of
% occurrences of the classical patterns $231$ and $213$ are the same on the
% set of permutations avoiding $132$,
% beside the pattern based statistics $231$ and $213$
% do not have the same distribution on this set.
% \emph{Almost avoidance} has also been considered;
% a permutation $\pi$ almost avoids a set $X$ of permutations
% if there is a way to remove a single element of $\pi$ to get a permutation
% that avoids all elements in $P$
% and $L_n(P)$ denotes the set of permutations of length $n$ that almost avoid
% $P$.
% It is shown in \cite{Griffiths:Smith:Warren:PMA:2011} that
% $L_n(213, 231) = L_n(132, 231) = L_n(132, 312) = L_n(213, 312)$.
This paper is organized as follows.
In Section~\ref{section:Definitions} the needed definitions are presented.
Section~\ref{section:both are (213,231)-avoiding} is devoted to presenting
an online linear-time algorithm in case both
permutations are wedge permutations,
whereas Section~\ref{section:sigma only avoid 231 and 213} focuses on the case
where only the pattern is a wedge permutation.
In Section~\ref{section:bivincular} we give a polynomial-time algorithm
for a bivincular wedge permutation pattern.
In Section~\ref{section:LCS} we consider the problem of finding the longest
wedge permutation pattern in permutations.

@@

When an element $\pi[\alpha]$ is smaller (resp. larger)
than an element $\pi[\beta]$  by position (in the order given by the permutation),
we say that $\pi_\alpha$ is on the left of (resp. on the right of) $\pi[\beta]$.
Besides, the element with the largest position is called the rightmost element
and the element with the smallest position is called the leftmost element.
For example, in the permutation $\pi = 51342$,
$5$ is on the left of $3$, the leftmost element is $5$
and the rightmost element is $2$.

>>>>

We focus in this paper on pattern matching issues for \emph{wedge permutations}
(\emph{i.e.}, those permutations that avoid both $213$ and $231$). 
%The number of $n$-wedge permutations is
%$t_0 = 1$ for $n = 0$ and
%$t_n =2^{n-1}$ for $n\geq 1$ \cite{Simion:Schmidt:EJC:1985},
%as they are in bijection with binary word of size $n-1$.
%On an individual basis,
%the permutations that do not have an occurrence of the permutation pattern $231$
%are exactly the \emph{stack-sortable permutations} and they are counted by
%the Catalan numbers \cite{Knuth:1997:ACP:260999}.
%	A stack-sortable permutation is a permutation whose elements may be sorted by
%	an algorithm whose internal storage is limited to a single stack data structure.
%	As for $213$, it is well-known that
% 	if $\pi = \pi_1\pi\,\ldots\,\pi_n$ avoids $132$, then its complement
% 	$\pi' = (n+1-\pi_1)(n+1-\pi_2)\,\ldots\,(n+1-\pi_n)$ avoids $312$, and
% 	the reverse of $\pi'$ avoids $213$.
% From a combinatorial point of view,
% B\'ona \cite{Bona:ElJC:2012}
% showed the rather surprising fact that the cumulative number of
% occurrences of the classical patterns $231$ and $213$ are the same on the
% set of permutations avoiding $132$,
% beside the pattern based statistics $231$ and $213$
% do not have the same distribution on this set.
% \emph{Almost avoidance} has also been considered;
% a permutation $\pi$ almost avoids a set $X$ of permutations
% if there is a way to remove a single element of $\pi$ to get a permutation
% that avoids all elements in $P$
% and $L_n(P)$ denotes the set of permutations of length $n$ that almost avoid
% $P$.
% It is shown in \cite{Griffiths:Smith:Warren:PMA:2011} that
% $L_n(213, 231) = L_n(132, 231) = L_n(132, 312) = L_n(213, 312)$.
This paper is organized as follows.
In Section~\ref{section:Definitions} the needed definitions are presented.
Section~\ref{section:both are (213,231)-avoiding} is devoted to presenting
an online linear-time algorithm in case both
permutations are wedge permutations,
whereas Section~\ref{section:sigma only avoid 231 and 213} focuses on the case
where only the pattern is a wedge permutation.
In Section~\ref{section:bivincular} we give a polynomial-time algorithm
for a bivincular wedge permutation pattern.
In Section~\ref{section:LCS} we consider the problem of finding the longest
wedge permutation pattern in permutations.

@@

When an element $\pi[\alpha]$ is smaller (resp. larger)
than an element $\pi[\beta]$  by position (in the order given by the permutation),
we say that $\pi[\alpha]$ is on the left of (resp. on the right of) $\pi[\beta]$.
Besides, the element with the largest position is called the rightmost element
and the element with the smallest position is called the leftmost element.
For example, in the permutation $\pi = 51342$,
$5$ is on the left of $3$, the leftmost element is $5$
and the rightmost element is $2$.
